\section{Introduction}

Data graphs are, essentially, relational structures in which nodes contain data values. This type of graph serves as the fundamental model underpinning two areas of great interest in knowledge representation and web data exchange: graph databases (see, e.g.,~\cite{AnglesG08,LV12}) and XML documents~\cite{xpath:w3c} (when the data graph has a tree structure). For this reason, a range of languages exists to formally describe and query data graphs. Two examples are Regular Queries~\cite{ReutterRV17}, a well-behaved fragment of Datalog, and XPath~\cite{xpath:w3c}, the most prominent approach for querying XML documents. The latter has received much attention recently, as it supports both the description of a data graph's topology and the manipulation of its data values~\cite{BojanczykMSS09}. Moreover, its semantics inspired modern query languages, such as the Graph Query Language (GQL) standard~\cite{FrancisGGLMMMPR23,FrancisGGLMMMPR23b}. Consequently, logical techniques developed for XPath may provide insights that extend beyond XML and are relevant to graph databases more broadly.


From a formal perspective,  XPath can be seen as a modal language~\cite{blackburn2001modal,blackburn06}. This makes it possible to define and use logical tools to reason about data graphs (see, e.g.,~\cite{ICDT14Jair,AbriolaDFF17,Areces2017a,Figueira18,AbriolaBFF18,ArecesF21,ArecesFC25}). The navigational part of XPath is much like standard modal logics, such as Propositional Dynamic Logic (PDL)~\cite{HKT00}, since it describes the interconnection between nodes in the data graph via so-called \emph{path expressions}. For instance, the path expression $((\ppath{a}\ppath{b}) \cup \ppath{c})$ describes a connection established either via the composition of an edge labelled by $\ppath{a}$ with one labelled by $\ppath{b}$, or via an edge labelled by $\ppath{c}$. Path expressions are used to build \emph{node expressions}.

Among node expressions, particularly interesting are those that describe \emph{data comparisons}. These have the form $\tup{\alpha =_\cmp \beta}$ and express that, starting from a given node of the graph, there exist a node $n$ reached via an $\alpha$-path and a node $m$ reached via a $\beta$-path (where $\alpha$ and $\beta$ are path expressions) such that the data value contained in $n$ equals the data value of $m$, for a given data comparison type $\cmp$. The modal perspective on XPath has already yielded important results, including complexity analyses, expressivity bounds, reasoning algorithms, and complete axiomatizations (see~\cite{ICDT14Jair,AbriolaDFF17,Areces2017a,Figueira18,AbriolaBFF18,ArecesF21} for further details).

% A notably underexplored direction is the use of XPath as a Description Logic (DL)~\cite{DLhandbook2003,DLbook17}, a natural perspective given the strong connection between Modal Logic and DL. The first steps in this direction were taken in~\cite{KostylevRV14,Kost:xpat15}, where XPath was investigated as an ontology language. In particular, the authors used DL-Lite reasoners to answer XPath queries. However, the fragment of XPath considered there includes only navigational features and leaves out data comparisons, which are a distinctive feature of the full language.

A notably underexplored direction is the interaction between XPath and Description Logics (DLs)~\cite{DLhandbook2003,DLbook17}, a natural perspective given the well-known correspondence between modal logics and DLs. Initial work in this direction was carried out in~\cite{KostylevRV14,Kost:xpat15}, where XPath query answering was studied in the presence of lightweight DL ontologies. In particular, the authors showed how simple DL reasoning techniques (like DL-Lite or EL) can be exploited to answer XPath queries over ontology-enriched data. However, the XPath fragments considered there are purely navigational and do not include data-comparison constructs, which constitute one of the distinguishing features of XPath.

In this article we aim to go beyond this limitation. Concretely, we focus on a fragment where path expressions are built from the composition of simple steps, setting aside tests and non-deterministic choice. For node expressions, however, we retain the full power of XPath with data equality and inequality. Additionally, we interpret path expressions as partial functions. We borrow notation from the modal logic $\knaltone$~\cite{blackburn2001modal}, which features $n$ modal operators whose respective accessibility relations are partial functions, and refer to this fragment $\xpath$. As we will discuss, these restrictions have an important computational effect, greatly reducing the complexity of various $\xpath$ reasoning tasks.

We consider that $\xpath$ remains expressive enough to model many interesting situations. In particular, consider the simplified graph database in~\Cref{fig:data-graph}. This database contains information about people, accounts, and banks, and adapts an example presented in~\cite{AbriolaBFF18}.

\begin{figure}[t]
    \begin{center}
       \begin{tikzpicture}[->, every node/.style={scale=0.825}]
       
        \node [state] (w1) at (0,0) {$\begin{array}{c}\mathsf{Person} \\  
            \text{\sf Id: 6365988}\end{array}$};      
        \node [state] (w2) at (3.75,0) {$\begin{array}{c}\mathsf{Person} \\  
            \text{\sf Id: 9001008}\end{array}$}; 
        \node [state] (w3) at (6,0) {$\begin{array}{c}\mathsf{Person} \\  
            \text{\sf Id: 3405616}\end{array}$}; 
        
        \node [state] (ww1) at (-1,-1.5){$\begin{array}{c}\mathsf{Account} \\  
            \text{\sf Id: 13209}\end{array}$}; 
        \node [state] (ww2) at (1,-1.5){$\begin{array}{c}\mathsf{Account} \\  
            \text{\sf Id: 004569}\end{array}$}; 
        \node [state] (ww3) at (3.75,-1.5){$\begin{array}{c}\mathsf{Account} \\  
            \text{\sf Id: 444900}\end{array}$}; 
        \node [state] (ww4) at (6,-1.5){$\begin{array}{c}\mathsf{Account} \\  
            \text{\sf Id: 712340}\end{array}$}; 
       
        \node [state] (www1) at (0,-3) {$\begin{array}{c}\mathsf{Bank} \\  
            \text{\sf Name: IB}\end{array}$}; 
        \node [state] (www2) at (3.75,-3) {$\begin{array}{c}\mathsf{Bank} \\  
            \text{\sf Name: NB}\end{array}$}; 

        \path (w1) edge node [label-edge, left] {$\mathsf{prim.acc}$} (ww1);
        \path (w1) edge node [label-edge, right] {$\mathsf{sec.acc}$} (ww2);
  
        \path (w2) edge node [label-edge, left] {$\mathsf{prim.acc}$} (ww3);
        \path (w2) edge node [label-edge, right] {$\mathsf{sec.acc}$} (ww4);
        \path (w3) edge node [label-edge, right] {$\mathsf{prim.acc}$} (ww4);

        \path (ww1) edge node [label-edge, left] {$\mathsf{in}$} (www1);
        \path (ww2) edge node [label-edge, above right] {$\mathsf{in}$} (www2);
        \path (ww3) edge node [label-edge, left] {$\mathsf{in}$} (www2);
        \path (ww4) edge node [label-edge, above left] {$\mathsf{in}$} (www2);
  
      \end{tikzpicture}
    \end{center}
    \caption{Example of a graph-structured database.} 
    \label{fig:data-graph}
    \medskip\medskip\medskip
\end{figure}

This database tells us that each person has an associated data value $\mathsf{Id}$, and that may own one or more accounts. The different accounts a person owns are captured as partial functions labelled \emph{primary account}, \emph{secondary account}, etc. Thus, a person's primary account is given by the functional relation $\mathsf{prim.acc}$, the secondary account is given by $\mathsf{sec.acc}$, and so on up to a fixed bound $n$ on the maximum number of accounts a person can have.

The database also captures the fact that bank accounts have corresponding $\mathsf{Id}$ data values. For instance, it tells us that the person whose $\mathsf{Id}$ is $\mathsf{6365988}$ has a primary account with $\mathsf{Id{:\ }13209}$, and a secondary account with $\mathsf{Id{:\ }004569}$. Moreover, it tells us that the account number $\mathsf{712340}$ is a joint account (for example, for married people). Bank accounts belong to a unique bank, identified by a data value $\mathsf{Name}$. In our example, there are two possible bank names: $\mathsf{IB}$ and $\mathsf{NB}$ (standing for International and National Bank, respectively).

For a given person, we can express interesting data comparisons of different kinds. An inequality comparison is exemplified by $[\mathsf{prim.account} \neq_{\mathsf{Id}} \mathsf{sec.acc}]$, which asks whether the primary and secondary accounts (if they exist) have different $\mathsf{Id}$ data values. An equality comparison, on the other hand, is illustrated by $\tup{\mathsf{prim.account\,in} =_{\mathsf{Name}} \mathsf{sec.acc\,in }}$, which states that a person has both a primary and a secondary account, and that both are in the same bank. Notice that the path expression $\mathsf{prim.account\,in}$ is interpreted as the composition of first accessing a node via the relation $\mathsf{prim.account}$, then accessing a node via the relation $\mathsf{in}$. In the example, the equality above holds for the person with $\mathsf{Id{:\ }9001008}$.

In addition to showing that $\xpath$ is expressive enough to capture interesting scenarios (particularly, with functional relations), the example above also serves to illustrate how this logic can be used in a DL setting. For instance, we can include definitions in a TBox and redefine long queries as short ones. To illustrate this point, a definition $\mathsf{elig} \doteq \tup{\mathsf{prim.account \, in} =_{\mathsf{Name}} \mathsf{sec.acc \, in }}$ could be used to define those persons that are eligible for certain bank benefits, whenever they have a primary and a secondary account in the same bank. With this definition, the new concept $\mathsf{elig}$ can be used as a succinct and more natural way to query the data graph, about such an eligibility criteria.
The data values illustrated above can be naturally accommodated in a DL setting resorting to \emph{concrete domains}~\cite{BaaderH91}. However, the complexity of DL with concrete domains is usually very high, so our goal is to investigate a formalism featuring data comparisons while still preserving good computational properties.

In this paper, we show that $\xpath$ can be viewed as a DL equipped with a very simple concrete domain for data values. This view allows us to obtain, as a byproduct, upper bounds for DL reasoning tasks in $\xpath$ by transferring existing results. It turns out, however, that these transferred bounds are not tight. Our main technical result is therefore to show that the problems of concept satisfiability, ABox consistency, and concept satisfiability w.r.t.\ a TBox are all \np-complete. Our proof strategy relies on selection functions, which require non trivial cases for manipulating data values. The obtained relatively low complexity bounds are potentially interesting for some applications, as these reasoning tasks have been proven to be at least \pspace-hard, and in many cases \exptime-hard or higher, for many DLs with simple concrete domains~\cite{Lutz1998-LUTDLW-2,Lutz02}.

% Data structures like the ones just described are usually handled in the DL literature via so-called concrete domains~\cite{BaaderH91}. However, it is known that concrete domains easily increment the computational complexity of DL reasoning tasks (see e.g.~\cite{Lutz02}). Thus, our first result is to show that $\xpath$ can be seen as a DL with a very simple concrete domain. This enables us to obtain upper bounds for DL reasoning tasks in $\xpath$ as a by product. It turns out that these bounds are not tight. Our main technical result is to show that the problems of concept satisfiability, ABox consistency and concept satisfiability w.r.t. a TBox are all \np-complete. These problems are very typical in DL, and our results show that $\xpath$ can be considered a DL with a very simple concrete domain featuring data comparisons, enjoying a good computational behaviour.

\paragraph*{Outline.} In~\Cref{sec:dxpath-f} we introduce $\xpath$, a fragment of XPath interpreted over models with functional accessibility relations. In~\Cref{sec:alcd} we present the DL with concrete domains $\alcd$ and recall relevant complexity results. In~\Cref{sec:xp-as-alcd} we define an equivalence-preserving translation from $\xpath$ to $\alcd$ (with~$\cdom$ a very simple concrete domain). This yields transference results, including complexity bounds for reasoning tasks. Bounds are refined in~\Cref{sec:complexity-xpath-f}, where we establish tight complexity for three typical DL problems over $\xpath$: concept satisfiability, ABox consistency, and concept satisfiability w.r.t.\ a TBox. We prove that all these problems are \np-complete. Our proof relies on a selection function to obtain a polynomial-size model property. In~\Cref{sec:final}, we offer some final remarks and future research directions.
