\section{Functional XPath with Data: $\xpath$} 
\label{sec:dxpath-f}

In this section we introduce a fragment of the language considered in, e.g.,~\cite{ICDT14Jair,KR16,AbriolaDFF17,AbriolaBFF18}, which we call $\xpath$. The fragment is inspired by the modal logic $\knaltone$ which has $n$ modal operators interpreted as partial functions~\cite{blackburn2001modal}. 
In brief, $\xpath$ is a two-sorted language which consists of  \emph{path} expressions and \emph{node} expressions.
%Path expressions capture the navigational fragment of the language, whereas node expressions capture properties.
Regarding path expressions, we restrict our attention to \emph{single steps} and \emph{composition}, excluding \emph{test} and \emph{choice}. Moreover, paths expressions will be interpreted as partial functions.  
We do not impose restrictions on node expressions.
As mentioned, these design decisions are aimed at establishing connections with some DL with concrete domains. 

In what follows, we assume that $\props = \set{p_1, \dots, p_n, p, q, r, \dots}$ is a countable set of \emph{propositional symbols}, $\mods=\set{\apath_1, \dots, \apath_n}$ is a finite set of \emph{modal labels}, and $\cmps = \set{\cmp_1, \dots, \cmp_m}$ is a finite set of \emph{comparison labels}. 

\begin{definition} [Language]\label{def:xpath:lang}
    The set $\pexp$ of path expressions (written $\alpha, \beta, \dots$)  and $\nexp$ of node expressions (written $\varphi, \psi, \dots$) of $\xpath$ are defined by the grammar:
    \begin{align*}
        \alpha, \beta & \Coloneqq
            \apath \mid
            \alpha \beta  \\
        \varphi, \psi & \Coloneqq
            p \mid
            \lnot \varphi \mid
            \varphi \land \psi \mid
            \tup{\alpha =_\cmp   \beta} \mid
            \tup{\alpha \neq_\cmp   \beta},
      \end{align*}
     where $p \in \props$, $\apath \in \mods$, $\cmp \in \cmps$. Other Boolean connectives are defined as usual. We also define $[\alpha=_\cmp\beta]$ as $\neg\tup{\alpha\neq_\cmp\beta}$ and $[\alpha\neq_\cmp\beta]$ as $\neg\tup{\alpha=_\cmp\beta}$. 
\end{definition}

Notice that path expressions are simply elements from $\mods^+$. This inductive definition in terms of composition will be handy in some proofs. Now we introduce the structures in which we evaluate $\xpath$ expressions.

\begin{definition}[Data Model] \label{def:model:xpath}
     A \emph{data model} is a structure defined as
     $\model = \tup{ \N, \D, \set{\R_{\apath}}_{\apath \in \mods}, \valuation,\set{\data_\cmp}_{\cmp \in \cmps} }$, where:
    \begin{enumerate}
        \item 
            $\N$ is a non-empty set of \emph{nodes};
        \item 
            $\D$ is a non-empty set of \emph{data values};
        \item 
            for each $\apath \in \mods$, $\R_{\apath}$ is a \emph{functional} relation on $\N$;
        \item 
            $\valuation: \props \to 2^\N$ is the valuation function;
        \item 
            $\data_\cmp:  \N \to \D$ is a function which assigns a data value to each node of the model with respect to $\cmp\in\cmps$.
    \end{enumerate}
    If $(n,m)\in\R_\apath$, sometimes we will use $\R_\apath(n)$ to denote $m$ (recall that $\R_\apath$ is functional) and we say that $\R_\apath(n)$ is \emph{defined}. If such an $m$ does not exist we say  $\R_\apath(n)$ is \emph{undefined}.
\end{definition}

Here, data models are a particular case of the data graphs from, e.g.,~\cite{ICDT14Jair}, in which accessibility relations are partial functions. 

In general, the set $\D$ of data values can be arbitrary, with the only condition imposed that it should be non-empty.  In what follows, and to properly establish the connection between XPath and DL with concrete domains, we will assume that a given $\D$ is fixed, containing all the data values that we need to consider, and the class of models is defined w.r.t.\ that fixed set of data values. 

%As discussed, this fragment is enough to describe some interesting situations. Moreover, we will show that it can be used as a description language enjoying nice computational properties.

\begin{definition}[Semantics] \label{def:xpath:semantics}
    Let $\model$ be a data model such that $\model=\tup{\N, \D, \set{\R_{\apath}}_{\apath \in \mods}, \valuation,\set{\data_\cmp}_{\cmp \in \cmps}}$. 
    We define the semantics of path and node expressions as:
    \[\begin{array}{l@{\, } c l}
        \truthset{ \apath }{\model}  
            & = & \R_\apath \\
        \truthset{ \alpha\beta }{\model}  
            & = & \{(m,n)\mid \text{ there is } l \text{ s.t. }  \\
            & & \qquad (m,l)\in\truthset{\alpha}{\model} \text{ and } (l,n)\in \truthset{\beta}{\model}\} \\
        \truthset{ p }{\model}  & = & \valuation(p)  \\
        \truthset{ \lnot \varphi }{\model}  
            & = & \N \setminus \truthset{ \varphi }{\model} \\
        \truthset{ \varphi \land \psi }{\model}  
            & = & \truthset{ \varphi  }{\model}  \cap \truthset{  \psi }{\model}
            \\    
        \truthset{ \tup{ \alpha =_\cmp \beta } }{\model}   & = & \{ m \mid  \text{there are } n,l \text{ s.t } \\
            & & \qquad (m,n) \in  \truthset{ \alpha }{\model} \text{ and } (m,l) \in  \truthset{ \beta }{\model}  \text{ and } \\
            & & \qquad \data_\cmp(n)=\data_\cmp(l) \} \\
        \truthset{ \tup{ \alpha \neq_\cmp \beta } }{\model}   & = & \{ m \mid  \text{there are } n,l \text{ s.t } \\
            & & \qquad (m,n) \in  \truthset{ \alpha }{\model} \text{ and } (m,l) \in  \truthset{ \beta }{\model}  \text{ and } \\
            & & \qquad \data_\cmp(n)\neq\data_\cmp(l) \}.
    \end{array}\]
    Note that the interpretation $\truthset{\alpha}{\model}$ of a path expression is a partial function on $\N$, whereas the interpretation $\truthset{\varphi}{\model}$ of a node expression is a subset of $\N$. 
    A node expression $\varphi$ is \emph{satisfiable} iff there is a data model $\model$ such that $\truthset{\varphi}{\model} \neq \emptyset$. Finally, we write $\model,n\models\varphi$ (resp. $\model,n,m\models\alpha$) iff $n\in\truthset{\varphi}{\model}$ (resp. $(n,m)\in\truthset{\alpha}{\model}$).
\end{definition}

        